Partitioning growth-dependent from growth-independent influences on tropical tree mortality

### 2018-08-22 ms/manuscript.Rnw
Incorporate Joe's comments

\section*{An additive hazards models with species effects}

@@

Here we propose the following hazard function, incorporating individual-, species-, and census-period effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate (our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$. The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.

@@

\begin{figure*}[!ht]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill and a logloss $\approx$ 0.693 indicating a model unable to distinguish an observed death from a survival.
A) Six sequential stages of model selection with each stage increasing in complexity. \textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\,\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate (i.e. carbon budget). This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters. Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
\label{fig2}
\end{figure*}

@@

In order to examine how species differed in hazard functions and to understand the effects of species traits, we used the entire dataset to fit a final model containing all three species traits plus a species random effect to capture inter-specific variation not explained by traits. The increase in predictive accuracy afforded by traits (stage 5) was fairly small relative to including species random effects on each parameter, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig. \ref{fig2} Stage 6).

@@

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in annual mortality probability with increasing growth rate. A) Estimated average curves for all 203 Barro Colorado Island species used in this study. B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values. B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8. Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e. the stochastic chance of death plus effect of low growth. Species with highest asymptote was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
\label{fig4}
\end{figure}

\subsection*{Explanatory power} 
The full model had an AUROC of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals. Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation in mortality observed among species.

Removal of the census random effect $\delta_t$ had little impact on either $r^2$ or AUROC. The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC. By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics. Both $r^2$ and AUROC dropped substantially when species random effects were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species. When both species traits and species random effects were removed, the resulting a model was by definition unable to capture any variation observed among species, and thus had low discriminatory ability. Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation. However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.

@@

Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}. Specifically, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates. By contrast, high maximum-dbh and high wood density species were found to be more tolerant of growth-independent hazards, corroborating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}. Our analyses also support other work \cite{vanGelder:2006exa, Kunstler:2016km} showing that higher wood density species tend also to be more tolerant of low growth rates and better able to tolerate shade and competition.

>>>>

\section*{An additive hazards model with species effects}

@@

Here we propose the following hazard function, incorporating individual-, species-, and census-period effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate ($X$; our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$. The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.

@@

\begin{figure*}[!ht]
\includegraphics{figures/fig2.pdf}
\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill and a logloss $\approx$ 0.693 indicating a model unable to distinguish an observed death from a survival.
A) Six sequential stages of model selection with each stage increasing in complexity. \textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\,\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate (i.e. carbon budget). This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters. Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square). Colours represent models that don't account for individual growth (white); use individual basal area growth (grey); or use diameter increment growth (black).}
\label{fig2}
\end{figure*}

@@

In order to examine how species differed in hazard functions and to understand the effects of species traits, we used the entire dataset to fit a final model containing all three species traits plus a species random effect to capture inter-specific variation not explained by traits. The increase in predictive accuracy afforded by traits (Fig. \ref{fig2}, compare best model at stage 2 with stage 5) was fairly small relative to including species random effects on each parameter (Fig. \ref{fig2} compare stages 5 and 6), signifying that substantial inter-specific variation in mortality rates remains unexplained.

@@

\begin{figure}[h!]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in annual mortality probability with increasing growth rate. A) Estimated average curves for all 203 Barro Colorado Island species used in this study. B,C,D) represent the average ($\pm$ 95\% credible intervals)  curve for a high (red) and low (blue) trait value. B) maximum dbh values 3 vs 180 cm; C) wood density values 0.2 vs 0.8 (g cm$^{-3}$); D) light demand values 0.5 vs 0.8. Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e. the stochastic chance of death plus effect of low growth. Species with highest asymptote in panel A was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
\label{fig4}
\end{figure}

\subsection*{Explanatory power} 
The full model had an Area Under the Roc Receiver Curve (AUROC) of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals. Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation in mortality observed among species.

Removal of the census random effect $\delta_t$ had little impact on either $r^2$ or AUROC. The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC. By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics. Both $r^2$ and AUROC dropped substantially when species random effects were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species. When both species traits and species random effects were removed, the resulting model was by definition unable to capture any variation observed among species, and thus had low discriminatory ability. Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation. However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.

@@

Our analyses support previous research that advocated that wood density, species maximum size and light demand are important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}. Specifically, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates. By contrast, high maximum-dbh and high wood density species were found to be more tolerant of growth-independent hazards, corroborating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}. Our analyses also support other work \cite{vanGelder:2006exa, Kunstler:2016km} showing that higher wood density species tend also to be more tolerant of low growth rates and better able to tolerate shade and competition.
